---
title: "Signature comparison between PROGENy and SPEED2"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de"
document: github_document
---

Recently, [SPEED2](https://github.com/molsysbio/speed2) was released upon the manuscript 
publication in NAR[ref](https://doi.org/10.1093/nar/gkaa236). They have increased pathway
coverage by mining an increased number of datasets. The full database could be accessed at
a TSV file in the repository 
( [`speed2_experiments.tsv`](https://github.com/molsysbio/speed2/blob/master/data-raw/speed2_experiments.tsv) ).
After further investigation of their repository, it seems that they provide a database with the signatures 
under a RData file at [`R/sysdata.rda`](https://github.com/molsysbio/speed2/blob/master/R/sysdata.rda).

## Setup environment
The environment will be set with a random seed number for reproducibility and an output folder
for processed data and figures.

### set env
```{r env}
options(stringsAsFactors = FALSE)
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./match_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```

### Load libraries
Essential libraries for R analysis.
```{r}
library(fgsea)
library(progeny)
library(ComplexHeatmap)
```

Register vars from env
```{r}
env_var <- ls()
```


## Data
### PROGENy signatures
We are going to use the full matrix of weights from PROGENy. For this, we use
the native function to get the matrix of weights
```{r}
progeny_model <- progeny::getModel("Human", top=Inf)
dim(progeny_model)
```
For the records, we show the pathway names
```{r}
print(colnames(progeny_model))
```

### SPEED2 signatures
Download the database of SPEED2 signatures if needed.
```{r}
fl <- "https://github.com/molsysbio/speed2/raw/master/R/sysdata.rda"
speed2_fl <- paste0(DATADIR,"/",basename(fl))
if(!file.exists(speed2_fl)) {
	download.file(url=fl, destfile=speed2_fl)
}
```

Load SPEED2 data
```{r}
load(speed2_fl)
```

Which objects contains that rda file?
```{r}
setdiff(ls(), env_var)
```

Taking into account that PROGENy data is just a matrix of gene symbols X pathways.
We are interested just in three files to make easy the matching with PROGENy:

* `speed2_signatures` contains the signatures as a long data.frame.
* `speed2_pathways` contains pathway IDs to pathway names.
* `df_entrezID` points out hgnc gene symbols to entrezID.

For the records, we show the pathway names from SPEED2:
```{r}
print(speed2_pathways$p_id)
```

We will create a PROGENy-like matrix for further comparisons. For this we have to reformat
the data.
```{r}
# speed2_signatures2 <- merge(x=speed2_signatures, y=speed2_pathways,
# 			    by.x="pathway_number", by.y="pathway_number",
# 			    all.x=TRUE)
# speed2_signatures3 <- merge(x=speed2_signatures2, y=df_entrezID,
# 			    by.x="g_id", by.y="g_id", all.x=TRUE)
```

Just to explore the database: it seems that not all gene universe is equally represented 
along the 16 pathways. Probably they have pre-selected genes based on subsets of the 
experiment database to build the model.
```{r}
(speed2_path_cnt <- length(unique(speed2_signatures$pathway_number)))
(speed2_gene_universe <- length(unique(speed2_signatures$g_id)))

# All gene universe is represented along the 16 pathways?
hist(table(speed2_signatures$g_id))
```

We build a PROGENy-like matrix for further comparisons.
```{r}
# First we define a dictionary of pathways
pathway_n2id <- setNames(speed2_pathways$p_id, 
			 as.character(speed2_pathways$pathway_number))
# Second we define the universe of genes
## If there is any duplicated, we are going to select the first entry in numerical entry
## because that means it was the first recorded with that name.
if(any(duplicated(df_entrezID$gene))) {
 cat("WARN: Selecting genes based on order for duplicated genes\n",
     file=stdout())
 df_entrezID <- df_entrezID[order(df_entrezID$g_id, decreasing=FALSE),]
 df_entrezID <- df_entrezID[!duplicated(df_entrezID$gene),]
 
 speed2_signatures <- speed2_signatures[speed2_signatures$g_id %in% df_entrezID$g_id, ]
}

# Finally, we define the gene universe
genes_entrez2hgnc <- setNames(df_entrezID$gene, 
			      as.character(df_entrezID$g_id))
# Speed2 weights: we define weights as -log10(pval) * sign(zscore)
speed2_signatures$weight <- -log10(speed2_signatures$qval) *
	sign(speed2_signatures$zrank_signed_mean)

# Empty matrix to store the model
SPEED2_model <- matrix(NA, nrow=length(genes_entrez2hgnc), ncol=length(pathway_n2id),
		       dimnames=list(genes_entrez2hgnc, pathway_n2id))
# We split by chunks of pathways the signatures
speed2_signatures2 <- split(speed2_signatures, speed2_signatures$pathway_number)

for(pid in names(speed2_signatures2)) {
	signat <- speed2_signatures2[[pid]]$weight
	names(signat) <- genes_entrez2hgnc[as.character(speed2_signatures2[[pid]]$g_id)]
	if(any(is.na(names(signat)))) {
		cat(paste0("WARN: it seems there are NA's in gene symbols from signatures...",
			   "n=",sum(is.na(names(signat))),".\n",
			   "Those will be removed.\n"),
		    file=stdout())
		signat <- signat[!is.na(names(signat))] # I found one case of 1 entrezID not described 
	}

	#Dimension
	genes <- names(signat) 
	path <- pathway_n2id[as.character(pid)]
	# Replace
	SPEED2_model[genes, path] <- signat 
}
```

## Looking for a common space
First we take a look which pathways are common between the two methods.
We will rename them to match each other. In summary:

* Estrogen, Hypoxia, JAK-STAT, p53, TGFb, TNFa, VEGF.
* Both present Wnt, but one is uppercase - we will change it accordingly.
* `PROGENy` presents two separates, MAPK and PI3K, while those are jointly built in `SPEED2`.
```{r}
colnames(SPEED2_model)
colnames(progeny_model)
```

We reformat `pathways` accordingly: WNT is uppercased, SPEED2's MAPK+PI3K is duplicated and 
renamed for further comparison.
```{r}
# Uppercase Wnt
colnames(SPEED2_model)[which(colnames(SPEED2_model)=="Wnt")] <- "WNT"

# Split and rename double MAPK+PI3K
SPEED2_model <- cbind(SPEED2_model, "PI3K"=SPEED2_model[,"MAPK+PI3K"])
colnames(SPEED2_model)[which(colnames(SPEED2_model)=="MAPK+PI3K")] <- "MAPK"
# Get common pathways
(common_paths <- intersect(colnames(SPEED2_model), colnames(progeny_model)))
```


What about the genes? Well, actually there is a large overlap! 21k genes. However,
many of these present NA's.
```{r}
nrow(SPEED2_model)
nrow(progeny_model)
length(intersect(rownames(SPEED2_model),rownames(progeny_model)))

# How many NA's per pathway
barplot(colSums(is.na(SPEED2_model)), horiz = TRUE, las=1)

# get common genes
common_genes <- intersect(rownames(SPEED2_model), rownames(progeny_model))
```

## Signature comparison
First we subset the data to have common dimension
```{r}
S2 <- SPEED2_model[common_genes, common_paths]
P <- progeny_model[common_genes, common_paths]
```

Pearson correlation shows a strong positive correlation between matched pairs between both models.
With the exception of WNT and VGF pathways.

```{r pearson_cor, fig.width=9, fig.height=7, dpi=300}
cormat <- cor(S2, P, method="pearson", use="pairwise.complete.obs")

corhp <- Heatmap(cormat, name="Pearson cor", 
	cluster_rows = FALSE, cluster_columns = FALSE,
	row_title = "SPEED2", column_title="PROGENy",
	row_names_side = "left", column_names_side="top",
	row_names_gp = gpar(fontsize=22), column_names_gp = gpar(fontsize=22),
	row_title_gp = gpar(fontsize=32), column_title_gp = gpar(fontsize=32),
	heatmap_legend_param = list("legend_direction"="horizontal",
				    "labels_gp"=gpar(fontsize=20),
				    "title_gp"=gpar(fontsize=26)
				    ),
	cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(sprintf("%.2f",cormat[i, j]), x, y, gp = gpar(fontsize=16))
    })
draw(corhp, heatmap_legend_side="bottom")
```

```{r spearman_cor, fig.width=9, fig.height=7, dpi=300}
cormat <- cor(S2, P, method="spearman", use="pairwise.complete.obs")

corhp <- Heatmap(cormat, name="Spearman cor", 
	cluster_rows = FALSE, cluster_columns = FALSE,
	row_title = "SPEED2", column_title="PROGENy",
	row_names_side = "left", column_names_side="top",
	row_names_gp = gpar(fontsize=22), column_names_gp = gpar(fontsize=22),
	row_title_gp = gpar(fontsize=32), column_title_gp = gpar(fontsize=32),
	heatmap_legend_param = list("legend_direction"="horizontal",
				    "labels_gp"=gpar(fontsize=20),
				    "title_gp"=gpar(fontsize=26)
				    ),
	cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(sprintf("%.2f",cormat[i, j]), x, y, gp = gpar(fontsize=16))
    })
draw(corhp, heatmap_legend_side="bottom")
```

```{r jaccard_idx, fig.width=9, fig.height=7, dpi=300}
jaccard_idx <- function(A, B) {
	idx <- length(intersect(A,B)) / length(unique(c(A,B)))
	return(idx)
}

S2_foot <- sapply(colnames(S2), function(pid) {
			  rnk <- S2[,pid]
			  names(rnk) <- paste0(rownames(S2),
					       ifelse(sign(rnk)>0,"_POS","_NEG")
					       )
			  rnk <- sort(abs(na.omit(rnk)),decreasing=TRUE)
			  return(head(names(rnk),100))
    }, simplify=FALSE)

P_foot <- sapply(colnames(P), function(pid) {
			  rnk <- P[,pid]
			  names(rnk) <- paste0(rownames(S2),
					       ifelse(sign(rnk)>0,"_POS","_NEG")
					       )
			  rnk <- sort(abs(na.omit(rnk)),decreasing=TRUE)
			  return(head(names(rnk),100))
    }, simplify=FALSE)


jacmat <- matrix(NA, nrow=ncol(S2), ncol=ncol(P), 
		  dimnames=list(colnames(S2), colnames(P)))
for(ridx in rownames(jacmat)) {
	A <- S2_foot[[ridx]]
	for(cidx in colnames(jacmat)) {
		B <- P_foot[[cidx]]
		jacmat[ridx, cidx] <- jaccard_idx(A,B)
	}
}

jachp <- Heatmap(jacmat, name="Jaccard Index",col=c("blue","yellow"), 
	cluster_rows = FALSE, cluster_columns = FALSE,
	row_title = "SPEED2 (100-gene footprints)", 
	column_title="PROGENy (100-gene footprints)",
	row_names_side = "left", column_names_side="top",
	row_names_gp = gpar(fontsize=22), column_names_gp = gpar(fontsize=22),
	row_title_gp = gpar(fontsize=30), column_title_gp = gpar(fontsize=30),
	heatmap_legend_param = list("legend_direction"="horizontal",
				    "labels_gp"=gpar(fontsize=20),
				    "title_gp"=gpar(fontsize=26)
				    ),
	cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(sprintf("%.2f",jacmat[i, j]), x, y, gp = gpar(fontsize=16))
    })
draw(jachp, heatmap_legend_side="bottom")
```

### Total Enrichment Score
```{r}
S2_UP <- sapply(colnames(S2), function(pid) {
		       names(sort(na.omit(setNames(S2[,pid],
						   rownames(S2))),
				  decreasing=TRUE))[1:100]
    }, simplify=FALSE)

P_UP <- sapply(colnames(P), function(pid) {
		       names(sort(na.omit(setNames(P[,pid],
						   rownames(P))),
				  decreasing=TRUE))[1:100]
    }, simplify=FALSE)

S2_DN <- sapply(colnames(S2), function(pid) {
		       names(sort(na.omit(setNames(S2[,pid],
						   rownames(S2))),
				  decreasing=FALSE))[1:100]
    }, simplify=FALSE)

P_DN <- sapply(colnames(P), function(pid) {
		       names(sort(na.omit(setNames(P[,pid],
						   rownames(P))),
				  decreasing=FALSE))[1:100]
    }, simplify=FALSE)


dummyNES_mat <- matrix(NA, nrow=ncol(S2), ncol=ncol(P), 
		  dimnames=list(colnames(S2), colnames(P)))

stopifnot(all(colnames(P) == colnames(S2)))
stopifnot(all(names(P_UP) == names(S2_UP)))
stopifnot(all(names(P_DN) == names(S2_DN)))

# First SPEED2 enrichments
S_diffNES_mat <- matrix(NA, nrow=ncol(S2), ncol=ncol(P), 
		  dimnames=list(colnames(S2), colnames(P)))
for(ridx in rownames(TES_mat)) {
	S_rnk <- sort(na.omit(setNames(S2[,ridx],rownames(S2))))
	S_nes_UP <- fgsea(P_UP, S_rnk)$NES
	S_nes_DN <- fgsea(P_DN, S_rnk)$NES
	S_diffNES <- S_nes_UP - S_nes_DN
	S_diffNES_mat[ridx,] <- S_diffNES
}

# # Second PROGENy enrichments over SPEED2
# P_diffNES_mat <- matrix(NA, nrow=ncol(S2), ncol=ncol(P), 
# 		  dimnames=list(colnames(S2), colnames(P)))
# for(cidx in colnames(TES_mat)) {
# 	P_rnk <- sort(na.omit(setNames(P[,cidx],rownames(P))))
# 	P_nes_UP <- fgsea(S2_UP, P_rnk)$NES
# 	P_nes_DN <- fgsea(S2_DN, P_rnk)$NES
# 	P_diffNES <- P_nes_UP - P_nes_DN
# 	P_diffNES_mat[, cidx] <- P_diffNES
# }
 
# Second PROGENy enrichments over SPEED2
P_diffNES_mat <- matrix(NA, nrow=ncol(P), ncol=ncol(S2), 
		  dimnames=list(colnames(P), colnames(S2)))
for(ridx in rownames(TES_mat)) {
	P_rnk <- sort(na.omit(setNames(P[,ridx],rownames(P))))
	P_nes_UP <- fgsea(S2_UP, P_rnk)$NES
	P_nes_DN <- fgsea(S2_DN, P_rnk)$NES
	P_diffNES <- P_nes_UP - P_nes_DN
	P_diffNES_mat[ridx, ] <- P_diffNES
}
# Average both
TESmat <- (S_diffNES_mat + P_diffNES_mat) / 2

# Visualization
teshp <- Heatmap(TESmat, name="Total Enrich. Score", 
	cluster_rows = FALSE, cluster_columns = FALSE,
	row_title = "SPEED2", 
	column_title="PROGENy",
	row_names_side = "left", column_names_side="top",
	row_names_gp = gpar(fontsize=22), column_names_gp = gpar(fontsize=22),
	row_title_gp = gpar(fontsize=30), column_title_gp = gpar(fontsize=30),
	heatmap_legend_param = list("legend_direction"="horizontal",
				    "labels_gp"=gpar(fontsize=20),
				    "title_gp"=gpar(fontsize=26)
				    ),
	cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(sprintf("%.2f",TESmat[i, j]), x, y, gp = gpar(fontsize=16))
    })
draw(teshp, heatmap_legend_side="bottom")

```

